Safe Transferable Regions

### 2016-10-07 paper/type-inference.tex
type inference updates

Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
inference algorithm.
The algorithm consists of the following steps:
\begin{enumerate}
 \item \emph{Region Parameterization}.
   The first step elaborates the input program by introducing \emph{formal region parameters}
   (for each class and method), and \emph{region variables} (representing yet undetermined
   \emph{actual region parameters}). We also introduce for each class and method, a
   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
   over the region parameters of that class/method.

@@

 \item If the solver succeeds, it returns a solution $\soln$ consisting of a
   pair of substitution functions $\substFn_\rho$ and $\substFn_\varphi$ for
   free region and predicate variables, respectively, introduced in step 1.
   We apply these substitutions to the elaborated program to produce the final program.
\end{enumerate}

\begin{figure}
\begin{codeml}
Infer ($p$) =
  let $q$ = IntroduceRegionParameters($p$) in
  let $C$ = GenerateConstraints($q$) in
  match (SolveConstraints($C$) with
  | None $\longrightarrow$ None
  | Some ($\soln$) $\longrightarrow$ Some ($q[\soln]$)
\end{codeml}

\caption{The type inference algorithm}
\label{fig:type-inference-algo}
\end{figure}

@@

\emph{Recursive Classes}.
The region parameters for a recursive class is computed in
a similar fashion, with the following difference: any recursive
field is ignored while instantiating region parameters for the fields of
the class, and the region parameters of the recursive class are computed

as before. We then do parameter instantiation for all recursive fields,
such that their region annotations (the actual region parameters) are
exactly the same as the (formal) region parameters of the class.
The following example illustrates this for a non-generic \C{List} class.
The resulting class represents a linked list with spine in the region
$\rho_0$ and data objects in the region $\rho_1$.

@@

\begin{theorem}
\emph{(\textbf{Soundness})}
For any $p \in \absof{\FB}$, if Infer($p$) returns Some($t$), then
(1) $\absof{t} = p$, and
(2) $t$ is well-typed.
\end{theorem}

However, in this context, completeness is more interesting than soundness.
(Note that an unsound inference algorithm can be made sound by simply type-checking
the resulting program.)
We believe that the converse of the above lemma can also be established:
any substitution for the region variables and the predicate variables that makes the
expression well-typed will also satisfy the generated constraints.
Thus, if there is any possible instantiation (of the region variables and predicate variables) of
the elaborated program produced by the first phase that type checks, then the generated
constraints are satisfiable, and the second phase will identify it.

However, the entire type inference algorithm is not complete, because of some heuristic
choices made during the first phase (elaboration phase), as explained below.

@@

\subsection{Proofs}

\input{typeinference-proofs}

>>>>

% Fig.~\ref{fig:type-inference-algo}
We now present a high-level outline of the type inference algorithm.
The algorithm consists of the following steps:
\begin{enumerate}
 \item \emph{Region Parameterization}.
   The first step elaborates the input program by introducing \emph{formal region parameters}
   (for each class and method), and \emph{region variables} (representing yet undetermined
   \emph{actual region parameters}). We also introduce for each class and method, a
   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
   over the region parameters of that class/method.

@@

 \item If the solver succeeds, it returns a solution $\soln$ consisting of a
   pair of substitution functions $\regionSubstFn$ and $\predSubstFn$ for
   free region and predicate variables, respectively, introduced in step 1.
   We apply these substitutions to the elaborated program to produce the final program.
\end{enumerate}

% \begin{figure}
% \begin{codeml}
% Infer ($p$) =
%   let $q$ = IntroduceRegionParameters($p$) in
%   let $C$ = GenerateConstraints($q$) in
%   match (SolveConstraints($C$) with
%   | None $\longrightarrow$ None
%   | Some ($\soln$) $\longrightarrow$ Some ($q[\soln]$)
% \end{codeml}
% 
% \caption{The type inference algorithm}
% \label{fig:type-inference-algo}
% \end{figure}

@@

\emph{Recursive Classes}.
The region parameters for a recursive class is computed in
a similar fashion, with the following difference: any recursive
field is ignored while instantiating region parameters for the fields of
the class, and the region parameters of the recursive class are computed
as before. We then do parameter instantiation for all recursive fields,
such that their region annotations (the actual region parameters) are
exactly the same as the (formal) region parameters of the class.
The following example illustrates this for a non-generic \C{List} class.
The resulting class represents a linked list with spine in the region
$\rho_0$ and data objects in the region $\rho_1$.

@@

\begin{theorem}
\emph{(\textbf{Soundness})}
For any $p \in \absof{\FB}$, if type inference succeeds, it returns a $t$ such that
% Infer($p$) returns Some($t$), then
(1) $\absof{t} = p$, and
(2) $t$ is well-typed.
\end{theorem}

% \item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.

However, in this context, completeness is more interesting than soundness.
(Note that an unsound inference algorithm can be made sound by simply type-checking
the resulting program.)
Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
show that the second and third steps of the type-inference algorithm are complete.
%% We believe that the converse of the above lemma can also be established:
%% any substitution for the region variables and the predicate variables that makes the
%% expression well-typed will also satisfy the generated constraints.
%% Thus, if there is any possible instantiation (of the region variables and predicate variables) of
%% the elaborated program produced by the first phase that type checks, then the generated
%% constraints are satisfiable, and the second phase will identify it.
%% 
The only sources of incompleteness in the type inference algorithm are some heuristic
choices made in the first step (the elaboration phase), as explained below.

@@

\subsection{Miscellaneous points and to-dos}

% \input{typeinference-proofs}

\begin{itemize}
  \item for function-calls: need basic typing from core language to introduce region parameters
  \item discuss computation of least-solution to constraints?
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item use of alpha-renaming, and its use to discharge not-in contraints;
also, incorporate this into first stage rather than second (in Fig. 7).
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
\end{itemize}

### 2016-10-07 paper/typeinference-proofs.tex
type inference updates

Invariants, other points, and to-dos:
\begin{itemize}
  \item discuss computation of least-solution to constraints?
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item unify notation for substitution
  \item substitution required in correctness theorem as well
  \item use of alpha-renaming, and its use to discharge not-in contraints;
also, incorporate this into first stage rather than second (in Fig. 7).
  \item Modify CSP definition to be same as preceding; we can define the $\Delta$s 
    as auxiliary notation
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Every predicate variable has a single domain-constraint.
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
  \item Proof by relating proof-trees with constraint-generation trees; each can
    be constructed from the other; inductive lemma is as earlier
\end{itemize}

\begin{theorem}
Let $p$, $q$, $r$ and $C$ denote the values of the corresponding
variables in an execution of the type inference algorithm.
\begin{enumerate}
\item $\absof{q}$ = $p$
\item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
\item If $\sigma$ is any solution to $C$, then $q[\sigma]$ is well-typed.
\item If there is any substitution $\sigma$ such that $q[\sigma]$ is well-typed, then
$C$ has a solution.
\item If SolveConstraints($C$) returns Some($s$), then $s$ is a solution to $C$.
\item If $C$ has a  solution, then SolveConstraints($C$) will return some solution.
\end{enumerate}
\end{theorem}

>>>>

% \begin{theorem}
% Let $p$, $q$, $r$ and $C$ denote the values of the corresponding
% variables in an execution of the type inference algorithm.
% \begin{enumerate}
% \item $\absof{q}$ = $p$
% \item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
% \item If $\sigma$ is any solution to $C$, then $q[\sigma]$ is well-typed.
% \item If there is any substitution $\sigma$ such that $q[\sigma]$ is well-typed, then
% $C$ has a solution.
% \item If SolveConstraints($C$) returns Some($s$), then $s$ is a solution to $C$.
% \item If $C$ has a  solution, then SolveConstraints($C$) will return some solution.
% \end{enumerate}
% \end{theorem}
